---
title: "Introduction to Exploratory Data Analysis"
subtitle: "Getting to Know Your Data with R"
author: "STEMINAR"
format:
  revealjs:
    theme: [default, ../custom.scss]
    slide-number: true
    transition: slide
    code-line-numbers: true
    highlight-style: github
    code-overflow: wrap
    incremental: true
    chalkboard: false
    width: 1280
    height: 720
    fig-width: 8
    fig-height: 4
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## Welcome! {.noincremental}

Today we will learn how to:

-   Load and inspect data in R
-   Clean and prepare data for analysis
-   Visualize distributions with histograms and bar charts
-   Compare groups using faceted plots
-   Explore relationships with scatter plots and linear models
-   Calculate and interpret summary statistics

## What is EDA? {.noincremental}

**Exploratory Data Analysis** is the process of using visualizations and summary statistics to better understand the structure, patterns, and relationships in your data.

. . .

> The goal is to **ask questions** and let the data guide your understanding.

## Our Toolkit {.noincremental}

We need two R packages:

::::: columns
::: {.column width="50%"}
### `tidyverse`

A collection of packages for data wrangling and visualization including:

-   `ggplot2` (plots)
-   `dplyr` (data manipulation)
-   `forcats` (factors)
:::

::: {.column width="50%"}
### `openintro`

A package from our textbook that includes:

-   Practice datasets
-   The `loan50` dataset we'll explore today
:::
:::::

## Loading Libraries

```{r}
#| echo: true
#| code-line-numbers: "1|2"
library(tidyverse)
library(openintro)
```

. . .

**Tip:** Always load your libraries at the top of your script!

# Getting to Know the Data {background-color="#006CB3"}

## The `loan50` Dataset {.noincremental}

-   Data from **Lending Club**, a peer-to-peer lending platform
-   Contains a sample of **50 loans**
-   Each row (observation) = one individual loan
-   Each column (variable) = a characteristic of that loan

## Inspecting the Data with `glimpse()`

```{r}
#| echo: true
#| code-line-numbers: "|1"
glimpse(loan50)
```

## Key Questions to Ask {.noincremental}

When you first encounter a dataset, always ask:

-   Where did this data come from?
-   What does each row represent? (observational unit)
-   How many observations are there?
-   How many variables are there?
-   What types of variables are present? (numerical vs. categorical)

. . .

**For `loan50`:** 50 observations, 18 variables from Lending Club.

# Cleaning the Data {background-color="#006CB3"}

## Step 1: Select Relevant Variables

We don't need all 18 variables. Let's narrow it down to 9:

```{r}
#| echo: true
#| code-line-numbers: "1|2|3"
loans <- loan50 |>
  select(loan_amount, interest_rate, term, grade,
         state, total_income, homeownership, loan_purpose, total_credit_limit)
```

. . .

**What happened?**

-   Created a new object called `loans`
-   Used the **pipe** (`|>`) to pass `loan50` into `select()`
-   Kept only the 9 columns we care about

## Step 2: Convert Factor Variables

Some variables are stored as `fct` (factor) — let's convert to character:

```{r}
#| echo: true
loans <- loans |>
  mutate_if(is.factor, as.character)
```

. . .

Now we're ready to explore!

## Our Cleaned Data

```{r}
#| echo: true
glimpse(loans)
```

# Exploring Loan Amounts {background-color="#006CB3"}

## Question: How big do loans tend to be?

To answer this, we'll use:

-   A **histogram** to visualize the distribution
-   **Summary statistics** to quantify the center and spread

## Building a Histogram

```{r}
#| echo: true
#| code-line-numbers: "1|2|3|4-8|9"
#| output-location: slide
loans |>
  ggplot(aes(x = loan_amount)) +
  geom_histogram() +
  labs(
    x = "Loan Amount ($)",
    y = "Count",
    title = "Distribution of Loan Amounts"
  ) +
  theme_minimal()
```

## Anatomy of a `ggplot` Call {.noincremental}

``` r
data |>                          # Start with data
  ggplot(aes(x = variable)) +   # Set up the canvas
  geom_histogram() +             # Add the geometry
  labs(x = "", y = "",           # Add labels
       title = "") +
  theme_minimal()                # Apply a clean theme
```

. . .

Each piece is a **layer** added with `+`

## Summary Statistics

```{r}
#| echo: true
#| code-line-numbers: "1|2|3-8"
loans |>
  filter(!is.na(loan_amount)) |>
  summarise(
    mean = mean(loan_amount),
    median = median(loan_amount),
    sd = sd(loan_amount),
    Q1 = quantile(loan_amount, 0.25),
    Q3 = quantile(loan_amount, 0.75)
  )
```

## Reading the Distribution {.noincremental}

When describing a distribution, always mention:

-   **Shape:** Is it symmetric, skewed left, or skewed right?
-   **Center:** What is the typical value? (mean and median)
-   **Spread:** How variable are the values? (standard deviation, IQR)
-   **Outliers:** Are there any unusual observations?

# Comparing Groups {background-color="#006CB3"}

## Question: Does loan size differ by homeownership?

We can split our histogram by a categorical variable using `facet_wrap()`.

## Faceted Histogram

```{r}
#| echo: true
#| code-line-numbers: "1-8|9"
#| output-location: slide
loans |>
  ggplot(aes(x = loan_amount)) +
  geom_histogram() +
  labs(
    x = "Loan Amount ($)",
    y = "Count",
    title = "Loan Amounts by Homeownership Status"
  ) +
  theme_minimal() +
  facet_wrap(~homeownership, nrow = 3)
```

## Grouped Summary Statistics

Add `group_by()` before `summarise()` to get stats per group:

```{r}
#| echo: true
#| code-line-numbers: "1-2|3"
loans |>
  filter(!is.na(loan_amount)) |>
  group_by(homeownership) |>
  summarise(
    mean = mean(loan_amount),
    median = median(loan_amount),
    sd = sd(loan_amount),
    Q1 = quantile(loan_amount, 0.25),
    Q3 = quantile(loan_amount, 0.75)
  )
```

## How Many in Each Group?

Use `count()` and `arrange()` to build a frequency table:

```{r}
#| echo: true
loans |>
  count(homeownership) |>
  arrange(desc(n))
```

## Adding Proportions

Use `mutate()` to calculate the proportion:

```{r}
#| echo: true
#| code-line-numbers: "1-3|4"
loans |>
  count(homeownership) |>
  arrange(desc(n)) |>
  mutate(proportion = n / sum(n))
```

# Exploring Relationships {background-color="#006CB3"}

## Question: Does income predict loan size?

We use a **scatter plot** when we have two numerical variables:

-   **Predictor** (x-axis): `total_income`
-   **Outcome** (y-axis): `loan_amount`

## Scatter Plot

```{r}
#| echo: true
#| code-line-numbers: "1|2|3|4-8|9"
#| output-location: slide
loans |>
  ggplot(aes(x = total_income, y = loan_amount)) +
  geom_point() +
  labs(
    x = "Total Income ($)",
    y = "Loan Amount ($)",
    title = "Total Income vs. Loan Amount"
  ) +
  theme_minimal()
```

## Adding a Linear Model

Use `geom_smooth(method = lm)` to overlay a regression line:

```{r}
#| echo: true
#| code-line-numbers: "1-3|4"
#| output-location: slide
loans |>
  ggplot(aes(x = total_income, y = loan_amount)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(
    x = "Total Income ($)",
    y = "Loan Amount ($)",
    title = "Total Income vs. Loan Amount with Linear Model"
  ) +
  theme_minimal()
```

## Getting the Equation

The `lm()` function gives us the **intercept** and **slope**:

```{r}
#| echo: true
lm(data = loans, formula = loan_amount ~ total_income)
```

. . .

The general form of the equation is:

$$\hat{y} = b_0 + b_1 \times x$$

## Correlation and R-Squared

```{r}
#| echo: true
#| code-line-numbers: "1|2-7|8-10"
loans |>
  summarise(
    mean_income = mean(total_income),
    sd_income = sd(total_income),
    mean_loan = mean(loan_amount),
    sd_loan = sd(loan_amount),
    r = cor(loan_amount, total_income)
  ) |>
  mutate(
    r_squared = r^2
  )
```

## Interpreting the Model {.noincremental}

**Slope** → Direction of the association

-   Positive slope = as x increases, y tends to increase
-   Negative slope = as x increases, y tends to decrease

. . .

**R-squared** → Strength of the association

-   Tells us the proportion of variability in y explained by x
-   Closer to 1 = stronger linear relationship
-   Closer to 0 = weaker linear relationship

# Categorical Data {background-color="#006CB3"}

## Question: Why are people getting loans?

For categorical data, we use **bar charts** and **frequency tables**.

## Bar Chart of Loan Purpose

```{r}
#| echo: true
#| code-line-numbers: "1|2|3|4-8|9"
#| output-location: slide
loans |>
  ggplot(aes(y = fct_rev(fct_infreq(loan_purpose)))) +
  geom_bar() +
  labs(
    x = "Count",
    y = "Loan Purpose",
    title = "Purposes of Loans in Our Sample"
  ) +
  theme_minimal()
```

## A Helpful Trick: `fct_infreq()` {.noincremental}

``` r
aes(y = fct_rev(fct_infreq(loan_purpose)))
```

-   `fct_infreq()` → orders the categories by frequency
-   `fct_rev()` → reverses the order so the most frequent is at the top

. . .

This makes bar charts **much easier to read**!

## Supporting Frequency Table

```{r}
#| echo: true
loans |>
  count(loan_purpose) |>
  arrange(desc(n))
```

# Key Takeaways {background-color="#006CB3"}

## The EDA Workflow {.noincremental}

1.  **Load** your libraries and data
2.  **Inspect** the data (`glimpse()`, `view()`)
3.  **Clean** the data (`select()`, `mutate()`)
4.  **Ask a question**
5.  **Visualize** (histogram, bar chart, scatter plot)
6.  **Summarize** (mean, median, sd, correlation)
7.  **Interpret** and communicate your findings
8.  **Repeat** with new questions!

## Functions We Learned {.noincremental}

::::: columns
::: {.column width="50%"}
### Data Wrangling

-   `glimpse()`
-   `select()`
-   `mutate()`
-   `filter()`
-   `group_by()`
-   `summarise()`
-   `count()`
-   `arrange()`
:::

::: {.column width="50%"}
### Visualization

-   `ggplot()` + `aes()`
-   `geom_histogram()`
-   `geom_bar()`
-   `geom_point()`
-   `geom_smooth()`
-   `facet_wrap()`
-   `labs()`
-   `theme_minimal()`
:::
:::::

## Now It's Your Turn! {.noincremental}

Open the tutorial file and work through each section.

Remember:

-   🔴 **Red boxes** = edit and/or run code
-   🔵 **Blue boxes** = answer a question

. . .

**Happy exploring!**